Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-16 ms.tex
D_k -> w_k; typos

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $D_k$ such that $D_k \ne D_j$ for $k \ne j$, and we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
$\ci$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Here we consider a large population of haploids and a trait determined by a single genetic locus with $n$ possible alleles $A_1, A_2, \ldots, A_n$ and corresponding fitness values $D_1, D_2, \ldots, D_n$.
The mutation rates $C_k$ of individuals with allele $A_k$ are potentially different;
specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of $A_k$, $C_k$, to depend on the identity of the allele $A_k$.

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation, $f'_k$ in the next generation is given by
\begin{equation}
f_k' = \big(1 - C_k\big) \frac{D_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{C_j \frac{D_j}{\bar{w}} f_j},
\label{eq:mutation_model_1}
\end{equation}
for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{D_k f_k}$ is the population mean fitness.

@@

Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Specifically, the fitness of individuals with $k$ mutant alleles is $D_k$,
And the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
or it is \emph{beneficial} with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
Then after selection and mutation $f'$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{D_0}{\bar{w}} f_0 + \beta C_{1} \frac{D_{1}}{\bar{w}} f_{1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{D_n}{\bar{w}} f_n + \delta C_{n-1} \frac{D_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{D_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{D_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{D_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.
Here $\bar w=\sum_{k=0}^nw_kf_n$ is the mean fitness.

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main digonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is irreducible.

% Migration model 1
\paragraph*{Migration model 1.}\label{sec:mig_model_1}

%ARE THERE OTHER MIGRATION MODELS (MODEL 1)?
In this case we consider a large population of haploids that occupy $n$ demes, $A_1, \ldots, A_n$.
Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition among individuals of all demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.

@@

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$,
and $u$ and $v$ be the corresponding left and right eigenvectors.
%TODO If $MCD$ is primitive, then
Then,
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

Then,
\begin{equation}\label{mfp_intermediate}
\begin{aligned}
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v = 
u\tr (0 - Z_k + M Z_k)D v = \\
-v_k u_k D_k + v_k D_k u\tr[M]_k = \\
v_k D_k (u\tr[M]_k - u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
u\tr \bar{w}^* = u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\bar{w}^*$ and the $k$
element of $u$:
\begin{equation}
u_k \bar{w}^* = (1-C_k) D_k u_k + C_k D_k u\tr [M]_k.
\end{equation}
Multiplying both sides by $v_k$ (which is positive if $D_k>0$) and rearranging, we get 
\begin{equation}
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k) = v_k D_k(u\tr [M]_k - u_k),
\end{equation}
which when substituted into Eq.~\ref{mfp_intermediate} yields:
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k).
\end{equation}
Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (\bar{w}^* - D_k),
\end{equation}
which completes the proof.
\end{proof}

@@

\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial C_k} = 
v_k D_k (u\tr[M]_k - u_k) = \\
D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occuring.

Given $0 < D_k \le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.

@@

We turn our attention to the case where the transition occurs from a subset $K$ of the types.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
If the transition rates are controlled by a threshold rule, then 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases}.
\end{equation}
If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)}.
\end{aligned}
\end{equation}
Thus
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

@@

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}). 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.

@@

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\Big(\lambda_1 - \frac{D_k}{\bar{w}^*}\Big).
\end{equation}
Evaluating this expression at $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
\begin{multline}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign\Big(1 - \frac{D_k}{\bar{w}^*}\Big) =  
sign(\bar{w}^* - D_k),
\end{multline}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

Note that if the modifier has the same effect on all individuals (i.e., $C_k = \alpha$ and $\tilde{C}_k = \alpha^*$ for all $k$) then $\cl_{ex}$ becomes
\begin{equation}
\cl_{ex} = ((1 - \alpha^*)I + \alpha^* M) D /\bar{w}^*,
\end{equation}  
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha^* < \alpha$, will be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above, we find
$$
\frac{\partial \bar{w}^*}{\partial \alpha} = 
\bar{w}^* - 1 \le 0,
$$
because $\bar{w}^* = \sum_{k=1}^{n}{v_k D_k}$ (see~\nameref{sec:AppA}), and $D_k \le 1$ for all $k$.

@@

\begin{lemma}
The normalizing factor $\bar{w}$ in Eq.~\ref{eq:model} is the population mean fitness
\begin{equation}
\bar{w} = \sum_k{f_k D_k}.
\end{equation}
\end{lemma}

\begin{proof}
By definition, the normalizing factor $\bar{w}$ ensures that $\sum_k{f'_k}=1$.
Denote $e = (1, ..., 1)$ the vector with ones at all positions and rewrite $e\tr f' = 1$.
Therefore, 
\begin{multline}\label{eq:mean_fitness}
\bar{w} = 
\bar{w} e\tr f' = 
e\tr (I - C + M C) D f = \\
(e\tr I - e\tr C + e\tr M C) D f = \\
(e\tr - e\tr C + e\tr C) D f = \\
e\tr D f = 
\diag{D} f = 
\sum_k{f_k D_k},
\end{multline}
using the fact that $e\tr M = e\tr$ and $e\tr I = e\tr$ because $M$ and $I$ are stochastic matrices.
\end{proof}

Note that this lemma also ensures that the stable population mean fitness $\bar{w}^*=\sum_k{v_k D_k}$ is the leading eigenvalue of $(I - C + M C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.

@@

% Appendix D: inner stability
\subsection*{Appendix D}\label{sec:AppD}
\begin{lemma}[Stability of the equilibrium $v$]
If $C$ and $D$ are positive diagonal matrices, and $M$ is block irreducible, then
the equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable. \end{lemma}
\begin{proof}
Let $A=(I-C+MC)D$ and denote the leading eigenvalue of $A$ as $\bar{w}^*$.

@@

Now, let $\norm{x} = \sum_{k=1}^{n}{x_k}$, so that $x_t \goesto x$ implies
\begin{equation}
\norm{x_t} \goesto \norm{x}
\label{eq:norm_limit}
\end{equation}
as $t \goesto \infty$.

Rewrite Eq.~\ref{eq:model} as 
\begin{equation}
f' = \frac{A f}{\norm{A f}} 
\label{eq:model_norm}
\end{equation}
Note that when applying eq.~\ref{eq:model_norm} $t$ times, the normalization can be applied once~\cite[Appendix~A]{Wakano2004}.

Therefore, for a positive frequency vector $f$, we can apply $f \goesto f'$ (Eq.~\ref{eq:model_norm}) $t$ times, and using eqs.~\ref{eq:perron_projection} and ~\ref{eq:norm_limit}, we have
\begin{multline}
\frac{A^t f}{\norm{A^t f}} = 
\frac{\frac{1}{(\bar{w}^*)^t} A^t f}{\frac{1}{(\bar{w}^*)^t} \norm{A^t f}} = 
\frac{\big(\frac{A}{\bar{w}^*}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\bar{w}^*}\big)^t \cdot f }} \goesto
\frac{\alpha v}{\norm{\alpha v}} = v,
\end{multline}
since $t \goesto \infty$, as $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
This completes the proof.
\end{proof}

>>>>

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $w_k$ such that $w_k \ne w_j$ for $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
$\ci$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Here we consider a large population of haploids and a trait determined by a single genetic locus with $n$ possible alleles $A_1, A_2, \ldots, A_n$ and corresponding fitness values $w_1, w_2, \ldots, w_n$.
The mutation rates $C_k$ of individuals with allele $A_k$ are potentially different;
specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of $A_k$, $C_k$, to depend on the identity of the allele $A_k$.

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation, $f'_k$ in the next generation is given by
\begin{equation}
f_k' = \big(1 - C_k\big) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{C_j \frac{w_j}{\bar{w}} f_j},
\label{eq:mutation_model_1}
\end{equation}
for all $k=1, 2, \ldots, n$, where $\bar{w} = \sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.

@@

Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 \ge w_1 \ge \ldots > w_n$),
and the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
or it is \emph{beneficial} with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation $f'$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.
Here $\bar{w}=\sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main diagonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is irreducible.

% Migration model
\paragraph*{Migration model.}\label{sec:mig_model}

In this case we consider a large population of haploids that occupy $n$ demes, $A_1, \ldots, A_n$.
Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
The fitness of individuals in deme $A_k$ is $w_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition among individuals of all demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.

@@

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$,
and $u$ and $v$ be the corresponding left and right eigenvectors.
%TODO If $MCD$ is primitive, then
Then,
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - w_k).
\end{equation}

@@

Then,
\begin{equation}\label{mfp_intermediate}
\begin{aligned}
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v = 
u\tr (0 - Z_k + M Z_k)D v = \\
-v_k u_k w_k + v_k w_k u\tr[M]_k = \\
v_k w_k (u\tr[M]_k - u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
u\tr \bar{w}^* = u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\bar{w}^*$ and the $k$
element of $u$:
\begin{equation}
u_k \bar{w}^* = (1-C_k) w_k u_k + C_k w_k u\tr [M]_k.
\end{equation}
Multiplying both sides by $v_k$ (which is positive if $w_k>0$) and rearranging, we get 
\begin{equation}
\frac{u_k v_k}{C_k} (\bar{w}^* - w_k) = v_k w_k (u\tr [M]_k - u_k),
\end{equation}
which when substituted into Eq.~\ref{mfp_intermediate} yields:
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - w_k).
\end{equation}
Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (\bar{w}^* - w_k),
\end{equation}
which completes the proof.
\end{proof}

@@

\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial C_k} = 
v_k w_k (u\tr[M]_k - u_k) = \\
w_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

Given $0 < w_k \le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.

@@

We turn our attention to the case where the transition occurs from a subset $K$ of the types.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
If the transition rates are controlled by a threshold rule, then 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases}.
\end{equation}
If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - w_k)}.
\end{aligned}
\end{equation}
Thus
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - w_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k w_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

@@

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}). 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.

@@

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - w_k).
\end{equation}

@@

Now, applying Lemma~\ref{lemma:mfp} (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\Big(\lambda_1 - \frac{w_k}{\bar{w}^*}\Big).
\end{equation}
Evaluating this expression at $\tilde{C}_k = C_k$, i.e. at $\lambda_1=1$, gives us
\begin{multline}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign\Big(1 - \frac{w_k}{\bar{w}^*}\Big) =  
sign(\bar{w}^* - w_k),
\end{multline}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

Note that if the modifier has the same effect on all individuals (i.e., $C_k = \alpha$ and $\tilde{C}_k = \alpha^*$ for all $k$) then $\cl_{ex}$ becomes
\begin{equation}
\cl_{ex} = ((1 - \alpha^*)I + \alpha^* M) D /\bar{w}^*,
\end{equation}  
and the \emph{unified reduction principle} is in effect~\cite[eqs. 65, 72]{Altenberg2017}.
Therefore, we can expect that with these modifiers, an allele $M$ that reduces the transition rate, i.e. $\alpha^* < \alpha$, will be favored by natural selection.
Indeed, setting $C = \alpha I$ in Eq.~\ref{eq:model_equilibrium} and proceeding as in the above, we find
$$
\frac{\partial \bar{w}^*}{\partial \alpha} = 
\bar{w}^* - 1 \le 0,
$$
because $\bar{w}^* = \sum_{k=1}^{n}{v_k w_k}$ (see~\nameref{sec:AppA}), and $w_k \le 1$ for all $k$.

@@

\begin{lemma}
The normalizing factor $\bar{w}$ in Eq.~\ref{eq:model} is the population mean fitness
\begin{equation}
\bar{w} = \sum_k{f_k w_k}.
\end{equation}
\end{lemma}

\begin{proof}
By definition, the normalizing factor $\bar{w}$ ensures that $\sum_k{f'_k}=1$.
Denote $e = (1, ..., 1)$ the vector with ones at all positions and rewrite $e\tr f' = 1$.
Therefore, 
\begin{multline}\label{eq:mean_fitness}
\bar{w} = 
\bar{w} e\tr f' = 
e\tr (I - C + M C) D f = \\
(e\tr I - e\tr C + e\tr M C) D f = \\
(e\tr - e\tr C + e\tr C) D f = \\
e\tr D f = 
\diag{D} f = 
\sum_k{f_k w_k},
\end{multline}
using the fact that $e\tr M = e\tr$ and $e\tr I = e\tr$ because $M$ and $I$ are stochastic matrices.
\end{proof}

Note that this lemma also ensures that the stable population mean fitness $\bar{w}^*=\sum_k{v_k w_k}$ is the leading eigenvalue of $(I - C + M C) D$; see Eq.~\ref{eq:model_equilibrium} and the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007}.

@@

% Appendix D: inner stability
\subsection*{Appendix D}\label{sec:AppD}

\begin{lemma}[Stability of the equilibrium $v$]
If $C$ and $D$ are positive diagonal matrices, and $M$ is block irreducible, 
then the equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable.
\end{lemma}

\begin{proof}
Let $A=(I-C+MC)D$ and denote the leading eigenvalue of $A$ as $\bar{w}^*$.

@@

Now, let $\norm{x} = \sum_{k=1}^{n}{x_k}$, so that $x_t \goesto x$ implies
\begin{equation}
\norm{x_t} \goesto \norm{x},
\label{eq:norm_limit}
\end{equation}
as $t \goesto \infty$ and rewrite Eq.~\ref{eq:model} as 
\begin{equation}
f' = \frac{A f}{\norm{A f}}. 
\label{eq:model_norm}
\end{equation}

Therefore, for a positive frequency vector $f$, we can apply $f \goesto f'$ (Eq.~\ref{eq:model_norm}) $t$ times, 
and as $f \goesto Af$ is linear, one normalization ($x \goesto x/\norm{x}$) suffices~\citep[see][Appendix~A]{Wakano2004}.
Using eqs.~\ref{eq:perron_projection} and ~\ref{eq:norm_limit}, we have
\begin{multline}
\frac{A^t f}{\norm{A^t f}} = 
\frac{\frac{1}{(\bar{w}^*)^t} A^t f}{\frac{1}{(\bar{w}^*)^t} \norm{A^t f}} = 
\frac{\big(\frac{A}{\bar{w}^*}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\bar{w}^*}\big)^t \cdot f }} \goesto
\frac{\alpha v}{\norm{\alpha v}} = v,
\end{multline}
since $t \goesto \infty$, as $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
This completes the proof.
\end{proof}
